---
title: "Exploratory Data Analysis for Opinion Labeling in News Articles"
subtitle: "W241, Experiments and Causality | Spring 2019 | Final Project"
author: "Authors: Craig Fujii, Tako Hisada, Cameron Kennedy"
output:
  pdf_document: default
  html_document: default
---

# Introduction

This notebook is designed to take the output of the data preparation, using the file `POSurveyData.csv` with potential outcomes from the raw survey results as its input data, and performing exploratory data analysis to descriptively and visually understand the results.

This notebook does not contain any inferential statistical analysis, as that is contained in a separate notebook.

The input of this notebook is as follows:

* **Input:**
    * **POSurveyData.csv**: The csv file that is the output of the data preparation. It contains the potential outcomes.

```{r, results='hide'} 
#Clear the environment
rm(list=ls())

# load packages 
library(tidyverse)
library(data.table)
library(ggplot2)
```

# Load Data

This block loads the data and performs some basic preparation.

```{r}
#Set data folder and filename
cstDataFolder <- './data/' #cst for 'constant'
#cstDataFile <- 'POSurveyData.csv'
cstDataFile <- 'POSurveyData.rds'

#Load response data
# dt <- fread(file=paste(cstDataFolder, cstDataFile, sep=''),
#             header=T
#             )

dt <- readRDS(file=paste(cstDataFolder, cstDataFile, sep=''))

#Inspect the data - Commented out to save space when printing
#dt[sample(.N,10)]
```

# Basic Counts

```{r}
#Commented out to save time/space
#summary(dt, maxsum=10) #maxsum: integer, indicating how many levels should be shown for factors.
```

# Duration

```{r}
#Convert seconds to minutes
dt[, Duration:=Duration/60]

#Group, then summarize
dtDur <- dt[, .(Duration=mean(Duration)), by=.(userID,Source)]
hist(dtDur$Duration, breaks=c(0:40))
summary(dtDur)


```

```{r}
dtDur[, .(mean=mean(Duration), med=median(Duration)), keyby=Source]
```

# Plotting Strategy

Look at our outcome variables for all level 1 possibilities of covariates.

* Tone
* Intensity
* Factualness

And these are covariates to explore:

* Gender
* Age
* Completion time (need to generate)
* Race
* Income
* Education
* Political Views
* Political Party
* Interest in Politics
* Number of Correct Political Battery Questions
* Reddit Usage
* Other Social Media Usage
* Article Topic
* Article Conservative/Liberal
* Flight Order

Key Covariates to ignore:

* Registered to Vote

```{r}
#Eliminate those who failed 1 or more attention checks
dt <- dt[AC_Num_Correct==4]
```

# Diverging Bar Plots

```{r}
### Percentage Response Charts

PlotDivBar <- function(varIn) {

  for (depVar in c('tone','intensity','fact')) {

    #Wrangle data into plot format
    dtPlot <- dt[, .N, keyby=.(newsOp,Treatment,demoVar=get(varIn),dv=get(depVar))]
    dtPlot[, pctCount:=N/sum(N), by=.(newsOp,Treatment,demoVar)]
    
    #Eliminate blanks
    dtPlot <- dtPlot[demoVar!='']
    
    #Create legend labels to include N-Size.
    dtN <- dtPlot[, .(N=sum(N)), keyby=.(demoVar)]
    dtN[, labelWithN:=paste(demoVar, '\n(N=', N, ')', sep='')]
    newLabels <- dtN$labelWithN
    names(newLabels) <- dtN$demoVar
    #return(dtN)
    
    if (depVar=='tone') {
      #Make Negatives for Left Categories
      dtPlot[dv=='01_Very Liberal', pctCount := -pctCount]
      dtPlot[dv=='02_Liberal', pctCount := -pctCount]
      dtPlot[dv=='03_Slightly Liberal', pctCount := -pctCount]
      
      #Split Center Category
      dtPlot[dv=='04_Neutral', pctCount := 0.5*pctCount] #Cut in half
      dtFlipCenter <- dtPlot[dv=='04_Neutral']
      dtFlipCenter[, pctCount := -pctCount]
      dtPlot <- rbind(dtPlot, dtFlipCenter)
      
      #Set plotting parameters
      lowerBound <- -1
      dvTitle <- 'Political Tone'
      legTitle <- 'Political Tone'
      colorPal <- 'RdBu'
      colorRev <- -1
    }
    
    if (depVar=='intensity') {
      # <no negative values required>
  
      #Set plotting parameters
      lowerBound <- 0
      dvTitle <- 'Political Intensity'
      legTitle <- 'Political\nIntensity'
      colorPal <- 'BuPu'
      colorRev <- -1
    }
    
    if (depVar=='fact') {
      #Make Negatives for Left Categories
      dtPlot[dv=='01_Mostly factual', pctCount := -pctCount]
      dtPlot[dv=='02_Somewhat factual', pctCount := -pctCount]
      
      #Set plotting parameters
      lowerBound <- -1
      dvTitle <- 'Factualness'
      legTitle <- 'Factualness'
      colorPal <- 'RdGy'
      colorRev <- -1
    }
    
    #Notes:
      #This requires making two bar plots, one for the pos + one for the neg, hence
        #the two geom_bar calls.
      #`fct_rev` from tidyverse is handy, used to reverse the order of factor levels
    
    for (articleType in c('News','Opinion')) {
      pa <- ggplot() +
        geom_bar(data=dtPlot[newsOp==articleType & pctCount>=0],
                 mapping=aes(x=Treatment, y=pctCount, fill=fct_rev(dv)),
                 color='black',
                 stat='identity',
                 position='stack',
                 width = 0.8) +
        geom_bar(data=dtPlot[newsOp==articleType & pctCount<0],
                 mapping=aes(x=Treatment, y=pctCount, fill=dv),
                 color='black',
                 stat='identity',
                 position='stack',
                 width = 0.8) +
        coord_flip() +
        facet_grid(demoVar~., switch='both', labeller=labeller(demoVar = newLabels)) +
        scale_y_continuous(labels = scales::percent, limits = c(lowerBound, 1)) +
        scale_fill_brewer(palette=colorPal, direction=colorRev, name=legTitle) +
        theme(panel.spacing = unit(0.5, 'lines'),
              strip.text.y = element_text(angle=180)
              ) +
        ggtitle(paste('Response Distribution: ', dvTitle, '\n',
                      articleType, ' Articles', ' - Grouped by ', varIn,
                          sep='')) +
        ylab('Percent of Respondents')
    
    print(pa)
    }
  }
  #return(dtPlot)
}

#Tasks
  #DONE Make this a function!
  #DONE x axis pctCount to friendly
  #DONE x axis numbers to percentages
  #DONE x axis fix scale
  #DONE Title (Title, News/Op, Variable)
  #DONE Legend Title
  #DONE Legend N sizes
  #DONE Reverse colors for tone (democrats should be blue, republicans red)
  #DONE Change color for Political Intensity (green to red ... or maybe shades of all 1 color)
  #DONE Change color for Factualness (green to red ... or maybe shades of all 1 color)
```

```{r}
#Overall Plot
PlotDivBar('Finished') #Proxy for 'Overall ATE'
PlotDivBar('Source')
```

```{r}
#A-plotting we well go!
PlotDivBar('topicName')
```

```{r}
PlotDivBar('age')
```

```{r}
PlotDivBar('ArticleTone')
PlotDivBar('ArticleClarity')
PlotDivBar('gender')
#PlotDivBar('race')
#PlotDivBar('income')
#PlotDivBar('education')
PlotDivBar('pol_views')
PlotDivBar('pol_party')
PlotDivBar('Int_politics')
#PlotDivBar('use_reddit')
#PlotDivBar('use_soc_media')
PlotDivBar('flightOrder')
```

# Plots for Presentation

```{r, fig.height=3, fig.width=5.5}
#Wrangle data into plot format
dtPlot <- dt[, .N, keyby=.(newsOp,Treatment,demoVar=get('Finished'),dv=get('fact'))]
dtPlot[, pctCount:=N/sum(N), by=.(newsOp,Treatment,demoVar)]

#Eliminate blanks
dtPlot <- dtPlot[demoVar!='']

#Create legend labels to include N-Size.
dtN <- dtPlot[, .(N=sum(N)), keyby=.(demoVar)]
dtN[, labelWithN:=paste(demoVar, '\n(N=', N, ')', sep='')]
newLabels <- dtN$labelWithN
names(newLabels) <- dtN$demoVar
#return(dtN)


#Make Negatives for Left Categories
dtPlot[dv=='01_Mostly factual', pctCount := -pctCount]
dtPlot[dv=='02_Somewhat factual', pctCount := -pctCount]

#Set plotting parameters
lowerBound <- -1
dvTitle <- 'Factualness'
legTitle <- 'Factualness'
colorPal <- 'RdGy'
colorRev <- -1


#Notes:
#This requires making two bar plots, one for the pos + one for the neg, hence
  #the two geom_bar calls.
#`fct_rev` from tidyverse is handy, used to reverse the order of factor levels

articleType <- 'Opinion'
pa <- ggplot() +
geom_bar(data=dtPlot[newsOp==articleType & pctCount>=0],
         mapping=aes(x=Treatment, y=pctCount, fill=fct_rev(dv)),
         color='black',
         stat='identity',
         position='stack',
         width = 0.8) +
geom_bar(data=dtPlot[newsOp==articleType & pctCount<0],
         mapping=aes(x=Treatment, y=pctCount, fill=dv),
         color='black',
         stat='identity',
         position='stack',
         width = 0.8) +
coord_flip() +
#facet_grid(demoVar~., switch='both', labeller=labeller(demoVar = newLabels)) +
scale_y_continuous(labels = scales::percent, limits = c(-0.25,0.9)) +
scale_fill_brewer(palette=colorPal, direction=colorRev, name=legTitle) +
theme(panel.spacing = unit(0.5, 'lines'),
      strip.text.y = element_text(angle=180),
      legend.position='bottom'
      ) +
ggtitle(paste('Response Distribution: Factualness', '\n',
              articleType, ' Articles (N=3,712)',
                  sep='')) +
ylab('Percent of Respondents')

pa <- pa + guides(fill=guide_legend(ncol=2,nrow=2,byrow=TRUE))

print(pa)
```

# ATE Bar Plots (no longer using, in favor of plots above)

```{r, fig.width=12}

# #Plot function
# PlotATE <- function(varIn) {
#   
#   #Wrap in for loop for tone, intensity, and factualness
#   for (depVar in c('tone','intensity','fact')) {
#     
#     #Wrangle data into plot format
#     dtPlot <- dt[, .N, keyby=.(newsOp,Treatment,demoVar=get(varIn),dv=get(depVar))]
#     dtPlot[, pctCount:=N/sum(N), by=.(newsOp,Treatment,demoVar)]
#     
#         #Eliminate blanks
#     dtPlot <- dtPlot[demoVar!='']
#     
#     #Create legend labels to include N-Size.
#     #Pattern: c("Female" = "F", "Male" = "M", "Othedr" = "O")
#     dtN <- dtPlot[, .(N=sum(N)), keyby=.(demoVar)]
#     dtN[, labelWithN:=paste(demoVar, '\n(N=', N, ')', sep='')]
#     newLabels <- dtN$labelWithN
#     names(newLabels) <- dtN$demoVar
#     
#     #Tall to wide (to set up subtracting ATEs)
#     dtPlot <- dcast(dtPlot,
#                     newsOp+demoVar+dv~Treatment,
#                     value.var='pctCount',
#                     fill=0
#                     )
#     
#     #Debugging-DELETE!!!!!!!!
#     #return(dtPlot)
#     
#     #Calculate ATE
#     dtPlot[, ATE:=Labeled-Unlabeled]
#     
#     #Rename demoVar and dv
#     setnames(dtPlot, old='demoVar', new=varIn)
#     setnames(dtPlot, old='dv', new=depVar)
#     
#     #Set Plot Title
#     if (depVar=='tone') {dvTitle <- 'Political Tone'}
#     if (depVar=='intensity') {dvTitle <- 'Political Intensity'}
#     if (depVar=='fact') {dvTitle <- 'Perceived Factualness'}
#     
#     #Generate plots, one for news and one for opinion
#     for (articleType in c('News','Opinion')) {
#       pa <- ggplot(dtPlot[newsOp==articleType], aes_string(y='ATE', x=varIn, fill=varIn)) +
#           geom_bar(position="dodge", stat="identity", color="black") +
#           facet_grid(cols=vars(get(depVar))) +
#           theme(axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.5),
#                 legend.position="bottom"
#                 ) +
#           scale_fill_brewer(palette="Blues", labels=newLabels) +
#           #scale_fill_discrete(labels=newLabels) +
#           ggtitle(paste('Average Treatment Effect: ', dvTitle, ', ', articleType, ' Articles\n', 
#                         'Grouped by ', varIn,
#                         sep=''))  
#       
#       print(pa)
#     }
#   }
#   
#   #Used for debugging
#   #return(dtPlot)
# }
# 
# #Debugging
# #dtDebug <- PlotATE('gender')
# ```
# 
# ```{r}
# #A-plotting we well go!
# PlotATE('Finished') #Proxy for 'Overall ATE'
# PlotATE('gender')
# PlotATE('age')
# PlotATE('race')
# PlotATE('income')
# PlotATE('education')
# PlotATE('pol_views')
# PlotATE('pol_party')
# PlotATE('Int_politics')
# PlotATE('use_reddit')
# PlotATE('use_soc_media')
# dt[, flightOrder:=as.factor(flightOrder)] #Plot function expects a factor
# PlotATE('flightOrder')
# PlotATE('topicName')

```

# Doodling

Everything from here below is just me playing with functions to learn them and explore.

```{r}
# #Key Y
# # tone
# # fact
# # intensity
# 
# 
# #Key X
# # treat
# # topicName
# # newsOp
# 
# #Counts of tone by treat and newsOp
# dcast(dt[, .N, keyby=.(treat, tone, newsOp)], 
#       tone~treat+newsOp,
#       value.var='N'
#       )
# 
# #Counts of fact by treat and newsOp
# dcast(dt[, .N, keyby=.(treat, fact, newsOp)], 
#       fact~treat+newsOp,
#       value.var='N'
#       )
# 
# #Counts of intensity by treat and newsOp
# dcast(dt[, .N, keyby=.(treat, intensity, newsOp)], 
#       intensity~treat+newsOp,
#       value.var='N'
#       )

```

```{r}
# #Plot ATE, treat vs. not treat, 1 for news, 1 for op
# 
# dtPlot <- dt[, .(count=.N), keyby=.(treat, tone)]
# 
# ggplot(data=dtPlot, aes(x=tone, y=count, col=factor(treat), group=factor(treat))) +
#    geom_line()+
#    geom_point()+
#    ggtitle(paste('Probability Mass Function\n','strSubtitle',sep='')) +
#    theme(axis.text.x = element_text(angle = 90, hjust = 1))

```


```{r}
# ### TESTING PLOTS, ETC.
# 
# #Get counts and percentages
# dtPlot <- dt[, .N, keyby=.(newsOp,Treatment,FL_4_DO,tone)]
# dtPlot[, pctCount:=N/sum(N), by=.(newsOp,Treatment,FL_4_DO)]
# #dtPlot
# 
# dtPlot <- dcast(dtPlot,
#                 newsOp+FL_4_DO+tone~Treatment,
#                 value.var='pctCount',
#                 fill=0
#                 )
# dtPlot[, ATE:=Labeled-Unlabeled]
# #dtPlot
# 
# ggplot(dtPlot, aes(y=ATE, x=tone, fill=FL_4_DO)) + 
#     geom_bar(position="dodge", stat="identity") +    
#     facet_wrap(~newsOp)
# 
# ggplot(dtPlot[newsOp=='News'], aes(y=ATE, x=FL_4_DO, fill=FL_4_DO)) + 
#     geom_bar(position="dodge", stat="identity") +    
#     facet_wrap(~tone)

```