---
title: "Statistical Analysis for Opinion Labeling in News Articles"
subtitle: "W241, Experiments and Causality | Spring 2019 | Final Project"
author: "Authors: Craig Fujii, Tako Hisada, Cameron Kennedy"
output:
  html_document: default
  pdf_document: default
classoption: landscape
---

# Introduction

This notebook runs the statistical analysis of the experiment, featuring the `polr` (proportional odds logistic regression) function to determine statistical significance of the survey results.

The input of this notebook is as follows:

* **Input:**
    * **POSurveyData.csv**: The csv file that is the output of the data preparation. It contains the potential outcomes.

```{r, results='hide'} 
#Clear the environment
rm(list=ls())

# load packages 
library(data.table)
library(MASS)
library(ggplot2)
library(dplyr)
library(stargazer)

#Set global variables
stargazerOutput <- 'text'
#stargazerOutput <- 'latex'
#stargazerOutput <- 'html'
```

# Load Data

This block loads the data and performs some basic preparation.

```{r}
#Set data folder and filename
cstDataFolder <- './data/' #cst for 'constant'
cstDataFile <- 'POSurveyData.rds'

dt <- readRDS(file=paste(cstDataFolder, cstDataFile, sep=''))

#Eliminate those who failed 1 or more attention checks
#Comment this line out to see all responses
dt <- dt[AC_Num_Correct==4]

#Inspect the data - Commented out to save space when printing
#dt[sample(.N,10)]
```

# Official Experiment Results

Prior to running the results, the research team agreed to conduct only two primary experiments:

  1. Opinion Articles, Factualness

  2. Opinion Articles, Intensity

The decision to not run more experiments was made on the basis of answering the original research question and also looking at the pilot results which showed little difference in outcomes among the covariates in the experiment.  

The next major section contains additional analyses that are 'fishing' for additional findings, but that are not considered among the official results.

## Experiments 1 and 2:  Intensity and Factualness, Opinion Articles

```{r, results='asis'}
### Opinion Models ###
newsOpValue <- 'Opinion'

### INTENSITY ###

#Dependent variable with treatment only
modIOSolo <- polr(intensity ~ treat, 
                  data=dt[newsOp==newsOpValue],
                  Hess=T)
#Grab t-value
modIOSolo_tValue <- coef(summary(modIOSolo))['treat1','t value']

#Calculate p-value
modIOSolo_pValue <- pt(abs(modIOSolo_tValue), nobs(modIOSolo), lower.tail = F) * 2

#DV with article clarity, tone, and MTurk Political View
modIOArtClarTone <- polr(intensity ~ treat + ArticleClarity + ArticleTone + MturkPolView,
                         data=dt[newsOp==newsOpValue],
                         Hess=T)

#DV with article topic
modIOTopic <- polr(intensity ~ treat + MturkPolView + topicName,
                   data=dt[newsOp==newsOpValue],
            Hess=T)

#Kitchen Sink (a bunch of covariates)
modIOKitchen <- polr(intensity ~ treat + ArticleClarity + ArticleTone + MturkPolView +
                       factor(userID) +
                       Int_politics + pol_party + age + gender + income +
                       use_reddit + use_soc_media,
                     data=dt[newsOp==newsOpValue],
                     Hess=T)


### FACTUALNESS ###

#Dependent variable with treatment only
modFOSolo <- polr(fact ~ treat, 
                  data=dt[newsOp==newsOpValue],
                  Hess=T)
#Grab t-value
modFOSolo_tValue <- coef(summary(modFOSolo))['treat1','t value']

#Calculate p-value
modFOSolo_pValue <- pt(abs(modFOSolo_tValue), nobs(modFOSolo), lower.tail = F) * 2

#DV with article clarity, tone, and MTurk Political View
modFOArtClarTone <- polr(fact ~ treat + ArticleClarity + ArticleTone + MturkPolView,
                         data=dt[newsOp==newsOpValue],
                         Hess=T)

#DV with article topic
modFOTopic <- polr(fact ~ treat + MturkPolView + topicName,
                   data=dt[newsOp==newsOpValue],
            Hess=T)

#Kitchen Sink (a bunch of covariates)
modFOKitchen <- polr(fact ~ treat + ArticleClarity + ArticleTone + MturkPolView +
                       factor(userID) +
                       Int_politics + pol_party + age + gender + income +
                       use_reddit + use_soc_media,
                     data=dt[newsOp==newsOpValue],
                     Hess=T)

#Compare models with stargazer
stargazer(modIOSolo, modIOArtClarTone, modIOTopic, modIOKitchen,
          modFOSolo, modFOArtClarTone, modFOTopic, modFOKitchen,
          type=stargazerOutput,
          omit = c('Int_politics','pol_party','gender','age','income',
                   'use_reddit','use_soc_media','userID'),
          add.lines = list(c('Data Subset','Opinion','Opinion','Opinion','Opinion'),
                           c('Fixed Interest in Politics','No','No','No','Yes'),
                           c('Fixed Surveyed Political Party','No','No','No','Yes'),
                           c('Fixed Age','No','No','No','Yes'),
                           c('Fixed Gender','No','No','No','Yes'),
                           c('Fixed Income','No','No','No','Yes'),
                           c('Fixed Reddit Usage','No','No','No','Yes'),
                           c('Fixed Social Media Usage','No','No','No','Yes'),
                           c('Fixed Subject (`userID`)','No','No','No','Yes')
                           )
          )
```

## Experiments 3 and 4:  Intensity and Factualness, News Articles

```{r, results='asis'}
### News Models ###
newsOpValue <- 'News'

### INTENSITY ###

#Dependent variable with treatment only
modIOSolo <- polr(intensity ~ treat, 
                  data=dt[newsOp==newsOpValue],
                  Hess=T)
#Grab t-value
modIOSolo_tValue <- coef(summary(modIOSolo))['treat1','t value']

#Calculate p-value
modIOSolo_pValue <- pt(abs(modIOSolo_tValue), nobs(modIOSolo), lower.tail = F) * 2

#DV with article clarity, tone, and MTurk Political View
modIOArtClarTone <- polr(intensity ~ treat + ArticleClarity + ArticleTone + MturkPolView,
                         data=dt[newsOp==newsOpValue],
                         Hess=T)

#DV with article topic
modIOTopic <- polr(intensity ~ treat + MturkPolView + topicName,
                   data=dt[newsOp==newsOpValue],
            Hess=T)

#Kitchen Sink (a bunch of covariates)
modIOKitchen <- polr(intensity ~ treat + ArticleClarity + ArticleTone + MturkPolView +
                       factor(userID) +
                       Int_politics + pol_party + age + gender + income +
                       use_reddit + use_soc_media,
                     data=dt[newsOp==newsOpValue],
                     Hess=T)


### FACTUALNESS ###

#Dependent variable with treatment only
modFOSolo <- polr(fact ~ treat, 
                  data=dt[newsOp==newsOpValue],
                  Hess=T)
#Grab t-value
modFOSolo_tValue <- coef(summary(modFOSolo))['treat1','t value']

#Calculate p-value
modFOSolo_pValue <- pt(abs(modFOSolo_tValue), nobs(modFOSolo), lower.tail = F) * 2

#DV with article clarity, tone, and MTurk Political View
modFOArtClarTone <- polr(fact ~ treat + ArticleClarity + ArticleTone + MturkPolView,
                         data=dt[newsOp==newsOpValue],
                         Hess=T)

#DV with article topic
modFOTopic <- polr(fact ~ treat + MturkPolView + topicName,
                   data=dt[newsOp==newsOpValue],
            Hess=T)

#Kitchen Sink (a bunch of covariates)
modFOKitchen <- polr(fact ~ treat + ArticleClarity + ArticleTone + MturkPolView +
                       factor(userID) +
                       Int_politics + pol_party + age + gender + income +
                       use_reddit + use_soc_media,
                     data=dt[newsOp==newsOpValue],
                     Hess=T)

#Compare models with stargazer
stargazer(modIOSolo, modIOArtClarTone, modIOTopic, modIOKitchen,
          modFOSolo, modFOArtClarTone, modFOTopic, modFOKitchen,
          type=stargazerOutput,
          omit = c('Int_politics','pol_party','gender','age','income',
                   'use_reddit','use_soc_media','userID'),
          add.lines = list(c('Data Subset','Opinion','Opinion','Opinion','Opinion'),
                           c('Fixed Interest in Politics','No','No','No','Yes'),
                           c('Fixed Surveyed Political Party','No','No','No','Yes'),
                           c('Fixed Age','No','No','No','Yes'),
                           c('Fixed Gender','No','No','No','Yes'),
                           c('Fixed Income','No','No','No','Yes'),
                           c('Fixed Reddit Usage','No','No','No','Yes'),
                           c('Fixed Social Media Usage','No','No','No','Yes'),
                           c('Fixed Subject (`userID`)','No','No','No','Yes')
                           )
          )
```

# Fishing! Exploring Relationships in Data Subsets

## Function to Compute Stats for Demographic Subsets

```{r, results='asis'}

HTEbyDemo <- function(demoIn, dvIn='fact') {
  print(paste0('DEMOGRAPHIC: ', demoIn))

  for(newsOpValue in levels(dt[, newsOp])) {
    print(newsOpValue)
  
    for (item in levels(dt[, factor(get(demoIn))])) {
      #Dependent variable with treatment only
      mod <- polr(get(dvIn) ~ treat,
                  data=dt[newsOp==newsOpValue & factor(get(demoIn))==item],
                  Hess=T)
      
      #Grab t-value, calculate p-value
      mod_tValue <- coef(summary(mod))['treat1','t value']
      #print(coef(summary(mod)))  #Verbose, good for debugging
      mod_pValue <- pt(abs(mod_tValue), nobs(mod), lower.tail = F) * 2
      print(paste0(newsOpValue, ' articles, Category: ', item, ' - p-value: ', mod_pValue))
      
      stargazer(mod, type=stargazerOutput, dep.var.labels = dvIn)
      print('')
      
    }
    print('')
  }
}

```

```{r, results='asis'}
#HTEbyDemo('topicName','intensity')
HTEbyDemo('age')
#HTEbyDemo('MturkPolView')
#HTEbyDemo('gender')
#HTEbyDemo('income')
#HTEbyDemo('Source')
#HTEbyDemo('pol_party')
```

## Looking at Interaction Terms

```{r}
#Releveling the age variable to set 60-69 as a reference
#because that group had the least ATE

#dt[, .N, keyby=age]

#dt <- within(dt, age <- relevel(age, ref = '60-69 years'))
dt[, age:=relevel(age, ref = '60-69 years')]
```

```{r, results='asis'}
### Opinion Models ###
newsOpValue <- 'Opinion'

#demoVar to try:
# ArticleClarity
# ArticleTone 
# MturkPolView
# Int_politics
# pol_party
# age
# gender
# income
# use_reddit
# use_soc_media

demoVar <- 'age'

#Dependent variable with treatment only
modFOSolo <- polr(fact ~ treat, 
            data=dt[newsOp==newsOpValue],
            Hess=T)

#Interaction Model 1
modFOInt <- polr(fact ~ treat*get(demoVar),
            data=dt[newsOp==newsOpValue],
            Hess=T)

#Compare models with stargazer
stargazer(modFOSolo, modFOInt,
          type=stargazerOutput
          # omit = c('Int_politics','Source','pol_party',
          #          'age','income','use_reddit',
          #          'Source','income','userID'),
          )
```
